COVID-19 and the 2020 US Presidential Election: Counties with Extra COVID-19 Deaths Showed Less Support For Trump

### 2023-12-19 outputs/paper/paper.qmd
Update the final paper

---
title: "An Observational Study to COVID-19 and 2020 US Presidential Election: Counties with High Deaths per Case Rate Showed Diminished Support For Trump"
title-block-banner: true
author: "Yiliu Cao"
thanks: "Code and data from this analysis are available at: https://github.com/yiliuc/covid_and_trump_loss.git"
date: "today"
date-format: "long"
abstract: "This study investigates the causal inference between COVID-19 and Donald Trump's loss during the 2020 US Federal Election using the data from MIT Election Data Science Lab and Johns Hopkins University CSSE. The main methodology used in this paper is propensity score matching with an exploratory analysis of the optimal treatment. The key finding suggests that the counties with a death per case rate exceeding the 0.4 quantile threshold show a reduced voting preference for Trump. This paper also conducts a counterfactual analysis based on the treatment effect, indicating that Trump might have been secured to re-elect if the disparities in death per case rates were addressed. Future research should refine the propensity score calculations by incorporating additional variables, such as the winning party in each county during the 2016 election."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(dplyr)
library(knitr)
library(ggplot2)
library(maps)
library(mapdata)
library(RColorBrewer)
library(gridExtra)
library(tools)
library(dplyr)
library(ggplot2)
```

@@

In early 2020, the United States encountered its first cases of COVID-19. The situation escalated rapidly, with reported cases reaching 10,000 by March 19 and soaring to 100,000 just eight days later [@WikiCOVIDUSA2023]. Amidst this unfolding crisis, the U.S. Presidential Election was underway, ending with Trump's defeat to Biden. By Election Day on November 3, there were more than 9 million reported cases and about 200,000 deaths in the U.S. When talking about Trump's loss, the public refers to his mishandling of "his greatest test" [@Greenblatt2021] and believes that his win in the 2016 Election was a historical accident [@Bryant2020]. Conversely, some arguments suggest that COVID-19 was not the sole or decisive factor in Trump's defeat; his limits as a political strategy and wasting his advantage on the economy also contributed to his loss. This paper will investigate the causal effect between COVID and Trump's loss, and if so, to what extent it influenced the 2020 election outcome.

Existing research on this subject provides varied perspectives. Baccini et al. [@Baccini2021CovidElection] suggested that Trump could win without COVID, but simply using the COVID-19 infection rate as the only factor contributing to his loss would be too naive. Socio-economic factors like high education attainment and race diversity are also unignorable. Besides, their research emphasized the significant impact of COVID-19 on those urban areas where there were no stay-at-home orders, particularly for the "swing" states. In addition, Noland & Zhang [@NolandZhang2021] suggested that the deaths per case is a more crucial metric than the infection rate when analyzing the impact of COVID-19 on voting for Trump. However, they highlight the challenge in assessing COVID-19's impact on voting, as we may not know when the voters made their decisions. In contrast, Clarke et al. [@Clarke2021CovidTrump] conducted actual surveys before and after the election, suggesting that COVID-19 impacts voting but not the dominant one. The U.S.'s highly polarized political landscape might be more significant. Similarly, Hart [@Hart2021CovidTrump] also conducted a survey to ask about people's attitudes toward Trump. All the participants show natural or negative possession. They also find that some social movements like "Murder of George Floyd" may impact voting.

While all the above research investigated the correlation between COVID-19 and voting for Trump, previous studies predominantly employed models to identify variables influencing Trump's vote share and perform counterfactual analyses to estimate his voting performance in the absence of COVID-19 or reduction in deaths. Instead of using that method, this study employs propensity score matching to find the treatment effect and analyze whether counties more severely affected by COVID-19 exhibit distinct voting patterns. However, given all counties have experienced COVID-19, this paper will first find the optimal treatment group(s). Finally, I will conduct a counterfactual analysis to see if Trump could be re-elected with the treatment effect eliminated.

This paper will have five parts. I will introduce the data used in this paper and present data summaries and visualization in the @sec-data part. After that, I will explain the methods in this paper and the corresponding results in the @sec-methods and @sec-results parts, respectively. All the results will be interpreted and discussed in the @sec-discussion part, and I will conclude with limitations and drawbacks in the @sec-conclusion part.

@@

The data in this paper is either downloaded directly or accessed via API. `R` [@R] will be the computer language used in the paper. Moreover, this paper will also use different R packages. The data is cleaned using the package `dplyr` [@dplyr], `stringr` [@stringr], `tidyverse` [@tidyverse], `janiotr` [@janitor], `tools` [@tools] and `tidyr` [@tidyr]. In addition to that, `ggplot2` [@ggplot2], `RColorBrewer` [@RColorBrewer], `maps` [@maps], `mapdata` [@mapdata], `gridExtra` [@gridExtra] and `cowplot` [@cowplot] will be used later to make tables and plot graphs. Besides, to perform Propensity Score Matching, this paper will also employ `Matching` [@Matching].

@@

This paper comprised five data sets from three different sources, each corresponding to different topics. The primary data is from the MIT Election Data Science Club [@MITElectionLab], which builds open online data collections of the U.S. Federal or Senate Election results from national to county levels. The data extracted is called "County Presidential Election Returns 2000-2020," [@MITElectionData2022] with about 70,000 rows containing the voting patterns for each candidate and party by county since 2000. Besides that, the data also indicate the types of voting, such as "EARLY VOTE" and "ELECTION DAY." To analyze the voting patterns for Donald J. Trump, I only filter the 2020 U.S. Federal Election data for all counties and parties.

@@

This paper also incorporates the socioeconomic data from the American Community Survey (ACS) [@acs], which is an online open-source database conducted by the U.S. Census Bureau, containing the various socioeconomic factors at different geographic levels. The data sets extracted from ACS are 2020 five-year estimates of DP02, DP03 and DP05, covering the social, economic and demographic characteristics at the county level. Since there are thousands of variables, to ensure simplicity, I will only use the variables that are found to be significant in predicting the COVID-19 mortality rate from my previous paper [@CaoCOVID19Repo]. The descriptions of all variables can be found on @tbl-descriptions.

After having five data sets, they are merged into one big data by county. Using the merged data, I calculate the percentage of votes for each party (candidate) in the 2020 election. In addition, I also create new dummy variables indicating the winning party in both elections for each county. Moreover, to accurately compare the COVID cases and deaths across all counties, I transform the number of cases and deaths to infection and mortality rate by 10,000 citizens in each county. Using the conclusion from Noland & Zhang's work [@NolandZhang2021], I also calculate each county's deaths per case rate. The final data consists of 3107 rows with 36 columns. All the essential variables are described in @tbl-descriptions.

@@

```{r}
#| echo: false
#| message: false
#| label: tbl-covid
#| tbl-cap: The summary of COVID cases and deaths as of Election Day.
data %>% 
  group_by(winning_party) %>% 
  summarise(n = n(),
            case = mean(cases),
            inf = mean(infrate),
            deaths = mean(deaths),
            mort = mean(mortrate),
            dpc = mean(dpc),
            income = mean(mean_household_income)) %>% 
  rename(`Winning Party` = winning_party,
         `Count` = n,
         `Average Cases` = case,
         `Infection Rate` = inf,
         `Average Deaths` = deaths,
         `Mortality Rate` = mort,
         `DPC` = dpc,
         `Income` = income) %>% 
  kable(booktabs = TRUE)
```

@@

```{r}
#| echo: false
#| message: false
#| fig.width: 8 
#| fig.height: 8
#| label: fig-map
#| fig-cap: The ratio of votes for the Republican and the infection rate per 100k in each county

@@

# Methods{#sec-meethods}

@@

Typically, the treatment should be defined before conducting the analysis, and this aims to reduce the risk of p-hacking [@Frost2023PHacking]. However, given the objective of this paper, which is to find whether COVID-19 influenced Trump's loss in the last Election, pre-defining the treatment is challenging as all U.S. counties were affected by COVID-19. Therefore, this study begins with exploratory analysis to identify the treatment group exhibiting the most significant differences in voting patterns for Trump. I will establish specific cutoffs for classifying counties as having 'high' or 'low' death per case rates (DPC). In addition, from @fig-dpc-income, low and high-income counties also seem to have different voting behaviours. Therefore, both DPC and income will be considered when choosing the optimal treatment.

@@

Randomized controlled trials (RCT) are regarded as the most ideal and the golden way to estimate the treatment effect [@Kuss2016PropensityScore]. The reason is that RCTs are the only way that guarantee the equal distribution of known and unknown parameters. Therefore, the outcomes between the treatment and control groups will not be confounded. However, conducting an RCT is nearly impractical and impossible in many scenarios. Back to the coronary artery bypass example [@Austin2011PropensityScore], it's not ethically feasible to randomly assign patients to new or traditional treatments. Hence, conducting an RCT is nearly impossible.

In such situations, implementing Propensity Score Matching (PSM) can be a more practical way to find the treatment effects. The propensity score is the probability of an observation that receives the treatment based on the existing covariates using logistic regression. Using propensity scores can reduce the dimension [@Zhao2021PropensityR], meaning that we can describe an observation to a single score instead of multiple covariates. In addition, another significant advantage of PSM is design separation [@Zhao2021PropensityR], meaning that PSM can separate the covariates balancing and effects estimation. This is especially beneficial as we can observe from @fig-insurance that the distribution of counties with relatively high DPC rates is imbalanced. Finally, we can match two observations having similar propensity scores but from treatment and control groups separately. Then, we will have many matched pairs and estimate the treatment effects. The matching process is called propensity score matching.

In this paper, the propensity scores are used to estimate how likely a county is to receive treatment using multiple linear logistic regression. The equation will look like:

```{=tex}
\begin{align*}
\log\left(\frac{P(T=1|X)}{1 - P(T=1|X)}\right) = \beta_0 + \beta_1 X_1 +\cdots +\beta_p X_p
\end{align*}
```
where:\
* $P(T=1|X)$ is the probability of a county receiving treatment\
* $X_j\:(1\leq j\leq p)$ are the covariates\
* $\beta_0$ is the intercept
* $\beta_j\:(1\leq j\leq p)$ is the corresponding coefficients of each covariate\

@@

The counterfactual analysis is particularly necessary for those "swing" states in 2020 [@WikiSwingState2023b]. In the 2016 Federal Election, one key factor in Trump's defeat of Hillary is that he won seven out of eleven swing states. However, he only won three in 2020. Besides, if Donald Trump were able to hold onto three swing states which are Georgia, Arizona and Wisconsin, where the average margin of Democrats is only about 0.3%, the result would have been a 269-269 electoral tie decided in the House of Representatives. The presidential election is left up to members of the House of Representatives, and Trump could win.

To find the voting patterns for Trump, especially for the "swing" states, I will follow the official election procedure and use the winners-takes-all rule [@WikiWinnerTakeAll2023c]. That said, I will re-calculate the votes for Trump at the county level and summarize by state level. Then, the party with the highest total votes will take all the electoral votes in this state. Eventually, I will calculate the total electoral votes for each party to see whether Trump could re-elect if the treatment effects diminish.

@@

So, a question may arise regarding whether we should include all the income and DPC in the settings of treatment. The answer is NO, and we should only include the DPC. From @fig-dpc, a consistent quadratic relationship between income and voting patterns for Trump is evident regardless of how the cutoff for defining high DPC is adjusted. This pattern demonstrates that poorer counties (in the first half of the income spectrum) may show a positive or a negative correlation with voting for Trump depending on the cutoff we choose. In contrast, more affluent counties will only exhibit a negative correlation no matter which cutoff we choose. This quadratic pattern can be verified from @fig-dpc-income, which illustrates a similar relationship between votes and income.

Conversely, @fig-income presents a different narrative. All regression lines, irrespective of the DPC cutoff, display a linear or nearly linear relationship. We should exclude income because the choice of cutoff will significantly change the treatment effects; DPC is more "stable" than income. We can verify this from @fig-income, where each line shows a negative slope, meaning that the choice of cutoff for the "High DPC" group will not change the general voting patterns for Trump. Hence, including income would introduce extra variability in the treatment effects and increase the probability of p-hacking. The @fig-treatment shows a simple visualization of why this is the case.

@@

In my previous paper, I already showed that the proportion of residents with at least a bachelor's degree, income, the proportion of people with private insurance and without health insurance, the ratio of males, ratio of black and black residents are statistically in predicting the probability of a county has a high mortality rate or not. In this paper, I will directly use the results and only consider these variables when predicting the propensity scores.

@@

library(dplyr)
data <- read.csv(here("outputs/data/merged_data.csv"))

data_te <- data %>% 
  mutate(high_dpc = ifelse(dpc > quantile(dpc, 0.4), 1, 0),
         high_income = ifelse(mean_household_income > quantile(mean_household_income, 0.9), 1, 0),
         treatment = high_dpc)
highinf_model <- glm(treatment ~ prop_higher_education  + pctile +
                       no_insurance + private_insurance + males +
                       white_pct + black_pct, family = binomial(),
                      data = data_te)
data_te$prop_score <- predict(highinf_model, type = "response")

library(Matching)
library(kableExtra)
# Propensity score matching
rr <- Match(Y = data_te$pct_vote_rep, Tr = data_te$treatment,
            X = data_te$prop_score, M = 1)

data.frame(estimate = -0.018342,
           p = 0.021404,
           on = 3115,
           ot = 1869,
           mo=1869,
           mou = 5415) %>% 
  rename(`Treatment Effect` = estimate,
         `P value` = p,
         `Obs.` = on,
         `Treat. Obs.` = ot,
         `Treat. Obs` = mo,
         `Treat. Obs. Unw.` = mou) %>% 
  kable(booktabs = TRUE) %>% 
  add_header_above(c(" " = 2,
                     "Original" = 2, "Matched" = 2))
```

@@

@@

```{r}
#| tbl-cap: The summary of counterfactual votes for the eleven swing staes in 2020
#| label: tbl-counter
electoral_votes_state <- read.csv(here("outputs/data/electoral_votes_state.csv"))
swing_states_2020 <- c("FL", "PA", "MI", "WI", "AZ", "NC", "GA", "IA", "NV", "MN", "NH")

@@

@@

## First discussion point {#sec-first-point}

## Second discussion point

## Third discussion point

## Weaknesses and next steps

>>>>

---
title: "COVID-19 and the 2020 US Presidential Election: Counties with Extra COVID-19 Deaths Showed Less Support For Trump"
title-block-banner: true
author: "Yiliu Cao"
thanks: "Code and data from this analysis are available at: https://github.com/yiliuc/covid_and_trump_loss"
date: "today"
date-format: "long"
abstract: "This study investigates the relationship between COVID-19 and Donald Trump's loss during the 2020 US Federal Election using the data from MIT Election Data Science Lab and Johns Hopkins University CSSE. The main methodology used in this paper is propensity score matching with an exploratory analysis of the optimal treatment. The key finding is that counties with a death per case rate exceeding the 0.4 quantile threshold show a reduced voting preference for Trump. This paper also conducts a counterfactual analysis based on the treatment effect, indicating that Trump might have been re-elected if the disparities of extra deaths were addressed. Future research should incorporate the neighborhood effects on voting between different counties, and validate the causal inference used."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(dplyr)
library(knitr)
library(ggplot2)
library(maps)
library(mapdata)
library(RColorBrewer)
library(gridExtra)
library(tools)
library(dplyr)
library(ggplot2)
library(Matching)
library(kableExtra)
```

@@

In early 2020, the United States encountered its first cases of COVID-19. The situation escalated rapidly, with reported cases reaching 10,000 by March 19 and soaring to 100,000 just eight days later [@WikiCOVIDUSA2023]. Amidst this unfolding crisis, the U.S. Presidential Election was underway, ending with Trump's defeat to Biden. By Election Day on November 3, there were more than nine million reported cases and about 200,000 deaths in the U.S. When talking about Trump's loss, the public refers to his mishandling of "his greatest test" [@Greenblatt2021] and believes that his win in the 2016 Election was a historical accident [@Bryant2020]. Conversely, some arguments suggest that COVID-19 was not the sole or decisive factor in Trump's defeat; his limits as a political strategy and wasting his advantage on the economy also contributed to his loss. This paper will investigate the causal effect between COVID and Trump's loss, and if so, to what extent it influenced the 2020 election outcome.

Existing research on this subject provides varied perspectives. Baccini et al. [@Baccini2021CovidElection] suggested that Trump could win without COVID, but simply using the COVID-19 infection rate as the only factor contributing to his loss would be too naive. Socio-economic factors like high education attainment and race diversity are also unignorable. Besides, their research emphasized the significant impact of COVID-19 on those urban areas where there were no stay-at-home orders, particularly for the "swing" states. In addition, Noland and Zhang (2021) suggested that the deaths per case is a more crucial metric than the infection rate when analyzing the impact of COVID-19 on voting for Trump. However, they highlight the challenge in assessing COVID-19's impact on voting, as we may not know when the voters made their decision. In contrast, Clarke, Stewart, and Ho (2021) conducted surveys before and after the election, suggesting that COVID-19 impacts voting but not the dominant one. The U.S.'s polarized political landscape might be more significant. Similarly, Hart (2021) also conducted a survey to ask about people's attitudes toward Trump. All the participants show natural or negative possession. They also find that some social movements like "Murder of George Floyd" may impact voting.

While all the above research investigated the correlation between COVID-19 and voting for Trump, previous studies predominantly employed models to identify variables influencing Trump's vote share and perform counterfactual analyses to estimate his voting performance in the absence of COVID-19 or reduction in deaths. Instead of using that method, this study employs propensity score matching to find the treatment effect and analyze whether counties more severely affected by COVID-19 exhibit distinct voting patterns. However, given all counties have experienced COVID-19, this paper will first find the optimal treatment group(s). Finally,wewill conduct a counterfactual analysis to see if Trump could be re-elected with the treatment effect eliminated.

This paper will have five parts.wewill introduce the data used in this paper and present data summaries and visualization in the @sec-data. After that,wewill explain the methods in this paper and the corresponding results in the @sec-methods and @sec-results, respectively. All the results will be interpreted and discussed in the @sec-discussion, andwewill conclude with limitations and drawbacks in the @sec-conclusion.

@@

The data in this paper is either downloaded directly or accessed via API. `R` [@R] will be the computer language used in the paper. The data is cleaned using the package `dplyr` [@dplyr], `stringr` [@stringr], `tidyverse` [@tidyverse], `janiotr` [@janitor], `tools` [@R] and `tidyr` [@tidyr]. In addition to that, `ggplot2` [@ggplot2], `RColorBrewer` [@RColorBrewer], `maps` [@maps], `mapdata` [@mapdata], `gridExtra` [@gridExtra] and `cowplot` [@cowplot] will be used later to make tables and plot graphs. To perform propensity score matching, this paper will also employ `Matching` [@Matching].

@@

This paper comprised five data sets from three different sources, each corresponding to different topics. The primary data is from the MIT Election Data Science Club [@MITElectionLab], which builds open online data collections of the U.S. Federal or Senate Election results from national to county levels. The data extracted is called "County Presidential Election Returns 2000-2020," [@MITElectionData2022] with about 70,000 rows containing the voting patterns for each candidate and party by county since 2000. Besides that, the data also indicate the types of voting, such as "EARLY VOTE" and "ELECTION DAY." To analyze the voting patterns for Trump,weonly filter the 2020 U.S. Federal Election data for all counties and parties.

@@

This paper also incorporates the socioeconomic data from the American Community Survey (ACS) [@acs], which is an online open-source database conducted by the U.S. Census Bureau, containing the various socioeconomic factors at different geographic levels. The data sets extracted from ACS are 2020 five-year estimates of DP02, DP03 and DP05, covering the social, economic and demographic characteristics at the county level. Since there are thousands of variables, to ensure simplicity,wewill only use the variables that are found to be significant in predicting the COVID-19 mortality rate from my previous paper [@CaoCOVID19Repo]. The descriptions of all variables can be found on @tbl-descriptions.

The five data sets are merged into one big data by county. Using the merged data,wecalculate the percentage of votes for each party (candidate) in the 2020 election. In addition,wealso create new dummy variables indicating the winning party in both elections for each county. Moreover, to accurately compare the COVID cases and deaths across all counties,wetransform the number of cases and deaths to infection and mortality rate by 10,000 citizens in each county. Using the conclusion from Noland & Zhang's work [@NolandZhang2021],wealso calculate each county's deaths per case rate. The final data consists of 3107 rows with 36 columns. All the essential variables are described in @tbl-descriptions.

@@

```{r}
#| echo: false
#| message: false
#| label: tbl-covid
#| tbl-cap: The summary of COVID cases and deaths as of Election Day.
data %>% 
  group_by(winning_party) %>% 
  summarise(n = n(),
            case = round(mean(cases), 0),
            inf = round(mean(infrate), 0),
            deaths = round(mean(deaths), 0),
            mort = round(mean(mortrate), 0),
            dpc = round(mean(dpc), 0),
            income = round(mean(mean_household_income), 0)) %>% 
  rename(`Winning Party` = winning_party,
         `Count` = n,
         `Average Cases` = case,
         `Infection Rate` = inf,
         `Average Deaths` = deaths,
         `Mortality Rate` = mort,
         `Death per Case (DPC)` = dpc,
         `Income` = income) %>% 
  kable(booktabs = TRUE)
```

@@

```{r}
#| echo: false
#| message: false
#| fig.width: 8 
#| fig.height: 8
#| label: fig-map
#| fig-cap: The ratio of Republican votes and the death per case (DPC) rate in each county

@@

# Methods{#sec-methods}

@@

Typically, the treatment should be defined before conducting the analysis, and this aims to reduce the risk of p-hacking [@Frost2023PHacking]. However, given the objective of this paper, which is to find whether COVID-19 influenced Trump's loss in the last Election, pre-defining the treatment is challenging as all U.S. counties were affected by COVID-19. Therefore, this study begins with exploratory analysis to identify the treatment group exhibiting the most significant differences in voting patterns for Trump.wewill establish specific cutoffs for classifying counties as having 'high' or 'low' death per case rates (DPC). In addition, from @fig-dpc-income, low and high-income counties also seem to have different voting behaviours. Therefore, both DPC and income will be considered when choosing the optimal treatment.

@@

Randomized controlled trials (RCT) are regarded as the most ideal way to estimate the treatment effect [@Kuss2016PropensityScore]. The reason is that RCTs are the only way that guarantee the equal distribution of known and unknown parameters. Therefore, the outcomes between the treatment and control groups will not be confounded. However, conducting an RCT is difficult in many scenarios. Back to the coronary artery bypass example [@Austin2011PropensityScore], it's not ethically feasible to randomly assign patients to new or traditional treatments. Hence, conducting an RCT is nearly impossible.

In such situations, implementing Propensity Score Matching (PSM) can be a more practical way to find the treatment effects. The propensity score is the probability an observation would have been treated based on the existing covariates using logistic regression. Using propensity scores reduces the dimension [@Zhao2021PropensityR], meaning that we can describe an observation to a single score instead of multiple covariates. Another significant advantage of PSM is design separation [@Zhao2021PropensityR], meaning that PSM can separate the covariates balancing and effects estimation. This is especially beneficial as we can observe from @fig-insurance that the distribution of counties with relatively high DPC rates is imbalanced. Finally, we can match two observations with similar propensity scores, but from treatment and control groups, separately. Then, we will have many matched pairs and estimate the treatment effect. The matching process is called propensity score matching.

In this paper, propensity scores are used to estimate how likely a county is to receive treatment using multiple linear logistic regression. The equation is:

$$
\begin{aligned}
\mbox{Pr}(y_i = 1) = & \text{logit}^{-1}(\beta_0 + \beta_1 \times \text{prop\_higher\_education}\\
& + \beta_2 \times \text{income\_pctile} + \beta_3 \times \text{no\_insurance} \\
& + \beta_4 \times \text{private\_insurance} + \beta_5 \times \text{males} \\
& + \beta_6 \times \text{white\_pct} + \beta_7 \times \text{black\_pct})
\end{aligned}
$$
where:

* $\mbox{Pr}(y_i = 1)$ is the probability that a county has a "High" DPC rate
* $\beta_0$ is the intercept
* $\beta_j\:(1\leq j\leq 7)$ is the corresponding coefficients for seven predictors

@@

The counterfactual analysis is particularly necessary for those "swing" states in 2020 [@WikiSwingState2023b]. In the 2016 Election, Trump won seven out of eleven swing states. However, he only won three in 2020. If Trump had been able to hold onto three swing states Georgia, Arizona and Wisconsin where the average margin of Democrats is only about 0.3%, the result would have been a 269-269 electoral tie. The presidential election is left up to members of the House of Representatives, and Trump could win.

To find the voting patterns for Trump, especially for the "swing" states, we will follow the official election procedure and use the winners-takes-all rule [@WikiWinnerTakeAll2023c]. That said, we will re-calculate the votes for Trump at the county level and summarize by state level. Then, the party with the highest total votes will take all the electoral votes in this state. Eventually, we will calculate the total electoral votes for each party to see whether Trump could re-elect if the treatment effects diminish.

@@

A question may arise regarding whether we should include all the income and DPC in the settings of treatment. We include only the DPC. Because, from @fig-dpc, a consistent quadratic relationship between income and voting patterns for Trump is evident regardless of how the cutoff for defining high DPC is adjusted. This pattern demonstrates that poorer counties (in the first half of the income spectrum) may show a positive or a negative correlation with voting for Trump depending on the cutoff we choose. In contrast, more affluent counties will only exhibit a negative correlation no matter which cutoff we choose. This quadratic pattern can be verified from @fig-dpc-income, which illustrates a similar relationship between votes and income.

Conversely, @fig-income presents a different narrative. All regression lines, irrespective of the DPC cutoff, display a linear or nearly linear relationship. We should exclude income because the choice of cutoff will significantly change the treatment effects; DPC is more "stable" than income. We can verify this from @fig-income, where each line shows a negative slope, meaning that the choice of cutoff for the "High DPC" group will not change the general voting patterns for Trump. Hence, including income would introduce extra variability in the treatment effects and increase the probability of p-hacking.

@@

Cao (2023) already showed that the proportion of residents with at least a bachelor's degree, income, the proportion of people with private insurance and without health insurance, the ratio of males, ratio of black and black residents are statistically in predicting the probability of a county has a high mortality rate or not. Here,we will directly use the results and only consider these variables when predicting the propensity scores.

@@

library(dplyr)
data_te <- data %>% 
  mutate(high_dpc = ifelse(dpc > quantile(dpc, 0.4), 1, 0),
         high_income = ifelse(mean_household_income > quantile(mean_household_income, 0.9), 1, 0),
         treatment = high_dpc)
highinf_model <- glm(treatment ~ prop_higher_education  + pctile +
                       no_insurance + private_insurance + males +
                       white_pct + black_pct, family = binomial(),
                      data = data_te)
data_te$prop_score <- predict(highinf_model, type = "response")
rr <- Match(Y = data_te$pct_vote_rep, Tr = data_te$treatment,
            X = data_te$prop_score, M = 1)
# summary(rr)
extract_match_summary <- function(rr) {
    summary_output <- capture.output(summary(rr))
    estimate <- as.numeric(sub("Estimate...\\s*(-?\\d+\\.?\\d*e?-?\\d*)", "\\1", summary_output[2]))
    p_value <- round(as.numeric(sub("p.val......\\s*(\\d+\\.?\\d*e?-?\\d*)", "\\1", summary_output[5])), 4)
    original_num_obs <- as.numeric(sub("Original number of observations..............\\s*(\\d+)", "\\1", summary_output[7]))
    original_num_treated <- as.numeric(sub("Original number of treated obs...............\\s*(\\d+)", "\\1", summary_output[8]))
    matched_num_obs <- as.numeric(sub("Matched number of observations...............\\s*(\\d+)", "\\1", summary_output[9]))
    matched_num_obs_unweighted <- as.numeric(sub("Matched number of observations  \\(unweighted\\).\\s*(\\d+)", "\\1", summary_output[10]))
    return(list(
        estimate = estimate,
        p = p_value,
        on = original_num_obs,
        ot = original_num_treated,
        mo = matched_num_obs,
        mou = matched_num_obs_unweighted
    ))
}

result <- extract_match_summary(rr)

data.frame(estimate = result$estimate,
           p = result$p,
           on = result$on,
           ot = result$ot,
           mo = result$mo,
           mou = result$mou) %>% 
  rename(`Treatment Effect` = estimate,
         `P value` = p,
         `Obs.` = on,
         `Treat. Obs.` = ot,
         `Treat. Obs` = mo,
         `Treat. Obs. Unw.` = mou) %>% 
  kable(booktabs = TRUE) %>% 
  add_header_above(c(" " = 2,
                     "Original" = 2, "Matched" = 2))
```

@@

# mb  <- MatchBalance(treatment ~ prop_higher_education  + pctile +
#                        no_insurance + private_insurance + males +
#                        white_pct + black_pct, data=data_te, match.out=rr, nboots=10)

@@

```{r}
#| tbl-cap: The summary of counterfactual votes for the eleven swing states in 2020
#| label: tbl-counter
electoral_votes_state <- read.csv(here("outputs/data/electoral_votes_state.csv"))
swing_states_2020 <- c("FL", "PA", "MI", "WI", "AZ", "NC", "GA", "IA", "NV", "MN", "NH")

@@

```{r}
#| include: false
library(dplyr)
data <- read.csv(here("outputs/data/merged_data.csv"))

data_te <- data %>% 
  filter(dpc > quantile(dpc, 0.4)) %>% 
  mutate(rep_won20 = ifelse(pct_vote_rep > pct_vote_demo, 1, 0),
         demo_won20 = ifelse(pct_vote_rep < pct_vote_demo, 1, 0)) %>% 
  mutate(treatment = rep_won20)

highinf_model <- glm(treatment ~ prop_higher_education  + pctile +
                       no_insurance + private_insurance + males +
                       white_pct + black_pct, family = binomial(),
                      data = data_te)
data_te$prop_score <- predict(highinf_model, type = "response")

library(Matching)
library(kableExtra)
# Propensity score matching
rr <- Match(Y = data_te$pct_vote_rep, Tr = data_te$treatment,
            X = data_te$prop_score, M = 1)
summary(rr)
```

@@

## Defining the treatment is the most crucial part in this study
In an observational study, especially for conducting the causal inference, the treatment should always be settled before the analysis as it is effective to reduce the risk of p-hacking. If we want to focus whether the COVID-19 has generally influenced the voting for Trump compared to 2016, then we can take the voting for each county on both 2016 and 2020 and the treatment would simply be whether this county had COVID-19. However, this paper aims to investigate whether the extra COVID-19 death during the pandemic has influenced the voting for Trump. As all the U.S. counties had experienced COVID-19, it is necessary to manually define the cutoff of "high" extra COVID-19 deaths. Therefore, this study starts with an exploratory analysis in finding the optimal treatment group.

From @tbl-covid and @fig-dpc-income, it seems that a county's COVID death per case rate (DPC) and income levels are the two factors that are most related to the voting for Trump. Therefore the treatment will only incorporate these two factors. However, even though the main objective in defining the cutoff is to maximize the difference between the treatment and control group, we can observe from @fig-dpc and @fig-income that the effect of income on the voting for Trump is highly dependent on the choice of cutoff as it appears a quadratic relationship. A quadratic pattern is not desired as different cutoffs will bring large differences on the treatment effects. The two groups may even show opposite relation (see @fig-treatment) or the same sign but with a high cutoff. However, unlike the income, DPC shows a general cubic relationship to vote of Trump and hence the choice of cutoff will not generate huge different treatment effects. We can verify this from @fig-income where no matter how the cutoff of "high" income changes, there will always have a linear or at least a closely linear relationship between them. The change of cutoff will not bring so dramatic change of treatment effects (see @fig-treatment). Furthermore, we can observe that the "high" and "low" DPC rate counties behave most differently with a cutoff at 0.4 quantile in @fig-income, therefore the optimal treatment group is counties with a DPC rate higher than the 0.4 quantile.

```{r}
#| label: tbl-income
# show TE with different income levels
generate_rr_summaries <- function(data, cutoffs, treatment_var) {
  rr_list <- list()
  for (cutoff in cutoffs) {
    data_te <- data %>%
      mutate(high_dpc = ifelse(dpc > quantile(dpc, cutoff), 1, 0),
             high_income = ifelse(mean_household_income > quantile(mean_household_income, cutoff), 1, 0))
    data_te$treatment <- data_te[[treatment_var]]
    if (treatment_var == "high_income"){
      highinf_model <- glm(treatment ~ prop_higher_education +
                           no_insurance + private_insurance + males +
                           white_pct + black_pct, family = binomial(), data = data_te)
    }
    else{
      highinf_model <- glm(treatment ~ prop_higher_education + pctile +
                           no_insurance + private_insurance + males +
                           white_pct + black_pct, family = binomial(), data = data_te)
    }
    data_te$prop_score <- predict(highinf_model, type = "response")
    rr <- Match(Y = data_te$pct_vote_rep, Tr = data_te$treatment, X = data_te$prop_score, M = 1)
    rr_list[[as.character(cutoff)]] <- rr
  }
  summaries <- list()

  for (cutoff in names(rr_list)) {
    summaries[[cutoff]] <- extract_match_summary(rr_list[[cutoff]])
  }

  return(summaries)
}
cutoffs <- seq(0.1, 0.9, by = 0.1)

summaries_income <- generate_rr_summaries(data, cutoffs, "high_income")
summaries_dpc <- generate_rr_summaries(data, cutoffs, "high_dpc")

extract_data_from_summaries <- function(summaries) {
  estimates_list <- list()
  p_values_list <- list()
  original_num_obs_list <- list()
  original_num_treated_list <- list()
  matched_num_obs_list <- list()
  matched_num_obs_unweighted_list <- list()

  # Extract variables for each summary into the respective lists
  for (cutoff in names(summaries)) {
      estimates_list[cutoff] <- summaries[[cutoff]]$estimate
      p_values_list[cutoff] <- summaries[[cutoff]]$p
      original_num_obs_list[cutoff] <- summaries[[cutoff]]$on
      original_num_treated_list[cutoff] <- summaries[[cutoff]]$ot
      matched_num_obs_list[cutoff] <- summaries[[cutoff]]$mo
      matched_num_obs_unweighted_list[cutoff] <- summaries[[cutoff]]$mou
  }

  # Create a data frame from the lists
  return(data.frame(
    cutoffs = c("0.1", "0.2", "0.3", "0.4", "0.5", "0.6", "0.7", "0.8", "0.9"),
    estimate = unname(unlist(estimates_list)),
    p = unname(unlist(p_values_list)),
    on = unname(unlist(original_num_obs_list)),
    ot = unname(unlist(original_num_treated_list)),
    mo = unname(unlist(matched_num_obs_list)),
    mou = unname(unlist(matched_num_obs_unweighted_list))
  ))
}

df_income <- extract_data_from_summaries(summaries_income)
df_dpc <- extract_data_from_summaries(summaries_dpc)
df_income %>% 
  rename(`Cutoffs` = cutoffs,
       `Treatment Effect` = estimate,
       `P value` = p,
       `Obs.` = on,
       `Treat. Obs.` = ot,
       `Treat. Obs` = mo,
       `Treat. Obs. Unw.` = mou) %>% 
kable(booktabs = TRUE) %>% 
add_header_above(c(" " = 3,
                   "Original" = 2, "Matched" = 2))
```
